### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updated block jacobi strategy section in spn equations

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.67777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 1.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10dspj}
\end{table}
It is readily apparent from the tabulated data the point Jacobi
preconditioning is insufficient as a preconditioning mechanism for
problems of order $SP_5$ or higher. In addition, as upscatter is
introduced into the system, a much larger spectral radius is observed
at lower $SP_N$ orders, signaling a need for a better preconditioning
strategy to both ensure and improve covergence for Monte Carlo
methods.

@@

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we instead
choose \textit{block Jacobi} preconditioning as a left preconditioner
for the system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/block_preconditioning}
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks. Right: Inversion is trivial and decoupled
      by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
The inversion of this preconditioner is trivial as shown on the right
side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
individually and combined to form the inverse. For high performance
implementations this has several attractive properties. First, the
blocks in the matrix come from the energy/angle discretization of the
transport equation. Each block on the diagonal is bound to a mesh
element in the system (note there are 8 blocks in each of the sparsity
patterns with a mesh of $2 \times 2 \times 2$) and therefore we expect
the matrix elements forming the block to be entirely local. Second,
these blocks are typically dense and nearly lower triangular for many
transport problems meaning that established dense matrix methods based
on LU-factorization and other schemes can be used for fast inversion.

>>>>

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.67777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 1.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10uspj}
\end{table}
It is readily apparent from the tabulated data the point Jacobi
preconditioning is insufficient as a preconditioning mechanism for
problems of order $SP_5$ or higher. In addition, as upscatter is
introduced into the system, a much larger spectral radius is observed
at lower $SP_N$ orders, signaling a need for a better preconditioning
strategy to both ensure and improve covergence for Monte Carlo
methods.

@@

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we instead
choose \textit{block Jacobi} preconditioning as a left preconditioner
for the system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_preconditioning.png}
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion is trivial
      and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
The inversion of this preconditioner is trivial as shown on the right
side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
individually and combined to form the inverse. For high performance
implementations this has several attractive properties. First, the
blocks in the matrix come from the energy/angle discretization of the
transport equation as given by Eq~(\ref{eq:A_block_matrix}). Each
block on the diagonal is bound to a mesh element in the system (note
there are 8 blocks on the diagonal in each of the sparsity patterns
with a mesh of $2 \times 2 \times 2$) and therefore we expect the
matrix elements forming the block to be entirely local. Second, these
blocks are typically dense and nearly lower triangular for many
transport problems meaning that established dense matrix methods based
on LU-factorization and other schemes can be used for fast inversion
\citep{lapack_citation}.
